Fusing Filters with Integer Linear Programming

### 2014-05-16 2014betterfusionforfilters/Main.tex
typo in abstract...
More desc to benchmarks

\begin{abstract}
The key to compiling functional, collection oriented array programs into efficient code is to minimise memory trafffic.
Simply fusing subsequent array operations into a single computation is not sufficient; we also need to cluster \emph{separate} traversals of the same array into a single traversal.
Previous work demonstrated how Integer Linear Programming (ILP) can be used to cluster the operators in a general data-flow graph into subgraphs, which can be individually fused.
However, these approaches can only handle operations which preserve the size of the array, thereby missing out on some optimisation opportunities.
This paper addresses this shortcoming by extending the ILP approach with support for size-changing operations, using an external ILP solver to find good clusterings.
\end{abstract}

@@

\section*{Acknowledgements}
Many thanks are due to
Manuel Chakravarty,
Robert Clifton-Everest,
Kai Engelhardt,
Bill Kroon,
Frederik M.\ Madsen,
Abdallah Saffidine,
Carter Schonwald,
and Jingling Xue
for enlightening discussions relating to this work.

>>>>

\begin{abstract}
The key to compiling functional, collection oriented array programs into efficient code is to minimise memory traffic.
Simply fusing subsequent array operations into a single computation is not sufficient; we also need to cluster \emph{separate} traversals of the same array into a single traversal.
Previous work demonstrated how Integer Linear Programming (ILP) can be used to cluster the operators in a general data-flow graph into subgraphs, which can be individually fused.
However, these approaches can only handle operations which preserve the size of the array, thereby missing out on some optimisation opportunities.
This paper addresses this shortcoming by extending the ILP approach with support for size-changing operations, using an external ILP solver to find good clusterings.
\end{abstract}

@@

\section*{Acknowledgements}
Many thanks are due to
Manuel Chakravarty,
Robert Clifton-Everest,
Erik de Castro Lopo,
Kai Engelhardt,
Bill Kroon,
Frederik M.\ Madsen,
Abdallah Saffidine,
Carter Schonwald,
and Jingling Xue
for enlightening discussions relating to this work.

### 2014-05-16 2014betterfusionforfilters/section/05-Benchmarks.tex
More desc to benchmarks

\begin{tabbing}
MMMMM   \= MMMMMMM \= M \= MMMMMMM \= M \= MMMMMMM \= \kill
Minimise   \> $25 \cdot x_{sum1, gts} + 1  \cdot x_{sum1,sum2} + 25 \cdot x_{sum1, ys2} +$ \\
           \> $25 \cdot x_{gts, sum2} + 25 \cdot x_{gts, ys1} + 1 \cdot x_{sum2, ys1} +$ \\
           \> $25 \cdot x_{ys1, ys2}  + 5  \cdot c_{gts} + 5 \cdot c_{ys1} + 5 \cdot c_{ys2} $\\
Subject to \\
    \> $-5 \cdot x_{sum1, gts}$  \> $\le$ \> $\pi_{gts} - \pi_{sum1}$  \> $\le$ \> $5 \cdot x_{sum1, gts}$  \\
    \> $-5 \cdot x_{sum1, sum2}$ \> $\le$ \> $\pi_{sum2} - \pi_{sum1}$ \> $\le$ \> $5 \cdot x_{sum1, sum2}$ \\
    \> $-5 \cdot x_{sum1, ys2 }$ \> $\le$ \> $\pi_{ys2 } - \pi_{sum1}$ \> $\le$ \> $5 \cdot x_{sum1, ys2 }$ \\
    \> $-5 \cdot x_{gts,  ys1 }$ \> $\le$ \> $\pi_{ys1 } - \pi_{gts }$ \> $\le$ \> $5 \cdot x_{gts, ys1  }$ \\
    \> $-5 \cdot x_{sum2, ys1 }$ \> $\le$ \> $\pi_{ys1 } - \pi_{sum2}$ \> $\le$ \> $5 \cdot x_{sum2, ys1 }$ \\
    \> $-5 \cdot x_{ys1, ys2  }$ \> $\le$ \> $\pi_{ys2 } - \pi_{ys1 }$ \> $\le$ \> $5 \cdot x_{ys1, ys2  }$ 
\\[1ex]
    \> $   x_{gts, sum2 }$ \> $\le$ \> $\pi_{sum2} - \pi_{gts }$ \> $\le$ \> $5 \cdot x_{gts, sum2 }$ 
\\[1ex]
    \>                     \>       \> $\pi_{sum1} < \pi_{ys1}$ \\
    \>                     \>       \> $\pi_{sum2} < \pi_{ys2}$
\\[1ex]
    \> $ x_{gts,sum2} $    \> $\le$ \> $c_{gts}$
\\[1ex]
    \> $x_{gts, sum2}$     \> $\le$ \> $x_{sum1, sum2}$ \\
    \> $x_{sum1,sum1}$     \> $\le$ \> $x_{sum1, sum2}$ \\
    \> $x_{sum1, gts}$     \> $\le$ \> $x_{sum1, sum2}$
\end{tabbing}

\noindent
One minimal solution to this is:
\begin{tabbing}
MMMMMMMMMMMMMMMMMMMMMMMMMM \= M \= \kill
$x_{sum1, gts},~ x_{sum1, sum1},~ x_{sum1, sum2},~ x_{gts, sum2},~ x_{ys1,  ys2}$
    \> $=$ \> $0$ \\
$x_{sum1, ys2},~ x_{gts, ys1},~   x_{sum2, ys1}$
    \> $=$ \> $1$ 
\\[1ex]
$\pi_{sum1},~ \pi_{gts },~ \pi_{sum2}$
    \> $=$ \> $0$ \\
$\pi_{ys1 },~ \pi_{ys2 }$
    \> $=$ \> $1$ 
\\[1ex]
$c_{gts},~ c_{ys1},~ c_{ys2}$           
    \> $=$ \> $0$
\end{tabbing}
This minimal solution is not unique, though in this case the only other minimal solutions use different $\pi$ values, and denote the same clustering. Looking at just the non-zero variables in the objective function, the value is $25 \cdot x_{sum1,ys2} + 25 \cdot x_{gts,ys1} + 1 \cdot x_{sum2, ys1} = 51$. For illustrative purposes, note that objective function could be reduced by setting $x_{sum1,ys2} = 0$ (fusing $sum1$ and $ys1$), but this conflicts with other constraints. Since $x_{sum1, sum2} = 0$, we require $\pi_{sum1} = \pi_{sum2}$, but also \mbox{$\pi_{sum2} < \pi_{ys2}$}. These constraints cannot be satisfied, so a clustering that fused $sum1$ and $ys2$ would not also permit $sum1$ and $sum2$ to be fused.

We will now compare the clustering produced by our system, with the one implied by stream fusion. As described in \cite{lippmeier2013flow}, stream fusion cannot fuse a produced array into multiple consumers, or fuse operators that are not in a producer-consumer relationship. The corresponding values of the $x_{ij}$ variables are:
\begin{tabbing}
MMMMMMMMMMMMMMMMMMMMMMMMMM \= M \= \kill
$x_{gts, sum2}$
    \> $=$ \> $0$ \\
$x_{sum1, gts}, x_{sum1, sum2}, x_{ys1,  ys2}, x_{sum1, ys2}, x_{gts, ys1 }, x_{sum2, ys1}$
    \> $=$ \> $1$
\end{tabbing}
We can force this clustering to be applied in our integer linear program by adding the above equations as new constraints. Solving the resulting program then yields:
\begin{tabbing}
MMMMMMMMMMMMMMMMMMMMMMMMMM \= M \= \kill
$\pi_{sum1}, \pi_{gts }, \pi_{sum2}$
    \> $=$ \> $0$ \\
$\pi_{ys1 }, \pi_{ys2 }$
    \> $=$ \> $1$ \\
$c_{gts}, c_{ys1}, c_{ys2}$           
    \> $=$ \> $0$
\end{tabbing}
Note that although nodes $sum1$ and $sum2$ have equal $\pi$ values, they are not fused. Conversely, if two nodes have different $\pi$ values then they are never fused. The corresponding value of the objective function is: $25 \cdot x_{sum1, gts} + 1 \cdot x_{sum1,sum2} + 25 \cdot x_{sum1, ys2} + 25 \cdot x_{gts, ys1} + 1 \cdot x_{sum2, ys1} + 25 \cdot x_{ys1, ys2} = 102$.

% -----------------------------------------------------------------------------
\subsection{Closest Points}
The closest points benchmark is a divide-and-conquer algorithm that, given an array of points, finds the closest pair of points.
A midpoint is found, and the points are filtered into those above the midpoint, and those below.
The closest points in the filtered arrays are then recursively computed, and the two results merged.
As the filters are operated on recursively, there are no opportunities for fusing the filters, making our clustering the same as Megiddo's.
Our clustering is able to generate both filtered arrays in a single loop, unlike stream fusion, which requires a separate loop for each.

% -----------------------------------------------------------------------------
\subsection{QuadTree}
The QuadTree benchmark recursively computes a 2-dimensional space partitioning tree of an array of points.
At each step, the array of points is filtered into four boxes.
As with closest points, there are no opportunities for fusing the filters, and our clustering is the same as Megiddo's.
Our clustering produces all four filtered results in a single loop, whereas stream fusion requires four loops.

% -----------------------------------------------------------------------------
\subsection{QuickHull}
QuickHull is another program that is worth mentioning, as it highlights some fusion opportunities with filters.
The core of the QuickHull algorithm is shown below: given a line and an array of points, the points are filtered to those above the line, and the farthest point above the line is also found.

@@

Stream fusion cannot fuse @pts'@ and @ma@ together, as @pts'@ is used multiple times and cannot be inlined into @ma@.
Megiddo cannot fuse them either, as their iteration sizes are different.
Our formulation is able to fuse these together.
If the @fold@ is changed to operate over @pts@ instead of @pts'@, Megiddo's formulation \emph{is} able to fuse the two.
However, a programmer unaware of fusion would expect the original version to be faster, because @pts'@ is smaller than @pts@.

>>>>

\begin{tabbing}
MMMMM   \= MMMMMMM \= M \= MMMMMMM \= M \= MMMMMMM \= \kill
Minimise   \> $25 \cdot x_{sum1, gts} + 1  \cdot x_{sum1,sum2} + 25 \cdot x_{sum1, ys2} +$ \\
           \> $25 \cdot x_{gts, sum2} + 25 \cdot x_{gts, ys1} + 1 \cdot x_{sum2, ys1} +$ \\
           \> $25 \cdot x_{ys1, ys2}  + 5  \cdot c_{gts} + 5 \cdot c_{ys1} + 5 \cdot c_{ys2} $
\\[1ex]
Subject to 
    \> $-5 \cdot x_{sum1, gts}$  \> $\le$ \> $\pi_{gts} - \pi_{sum1}$  \> $\le$ \> $5 \cdot x_{sum1, gts}$  \\
    \> $-5 \cdot x_{sum1, sum2}$ \> $\le$ \> $\pi_{sum2} - \pi_{sum1}$ \> $\le$ \> $5 \cdot x_{sum1, sum2}$ \\
    \> $-5 \cdot x_{sum1, ys2 }$ \> $\le$ \> $\pi_{ys2 } - \pi_{sum1}$ \> $\le$ \> $5 \cdot x_{sum1, ys2 }$ \\
    \> $-5 \cdot x_{gts,  ys1 }$ \> $\le$ \> $\pi_{ys1 } - \pi_{gts }$ \> $\le$ \> $5 \cdot x_{gts, ys1  }$ \\
    \> $-5 \cdot x_{sum2, ys1 }$ \> $\le$ \> $\pi_{ys1 } - \pi_{sum2}$ \> $\le$ \> $5 \cdot x_{sum2, ys1 }$ \\
    \> $-5 \cdot x_{ys1, ys2  }$ \> $\le$ \> $\pi_{ys2 } - \pi_{ys1 }$ \> $\le$ \> $5 \cdot x_{ys1, ys2  }$ 
\\[1ex]
    \> $   x_{gts, sum2 }$ \> $\le$ \> $\pi_{sum2} - \pi_{gts }$ \> $\le$ \> $5 \cdot x_{gts, sum2 }$ 
\\[1ex]
    \>                     \>       \> $\pi_{sum1} < \pi_{ys1}$ \\
    \>                     \>       \> $\pi_{sum2} < \pi_{ys2}$
\\[1ex]
    \> $ x_{gts,sum2} $    \> $\le$ \> $c_{gts}$
\\[1ex]
    \> $x_{gts, sum2}$     \> $\le$ \> $x_{sum1, sum2}$ \\
    \> $x_{sum1,sum1}$     \> $\le$ \> $x_{sum1, sum2}$ \\
    \> $x_{sum1, gts}$     \> $\le$ \> $x_{sum1, sum2}$
\end{tabbing}
\noindent
One minimal solution to this is:
\begin{tabbing}
MMMMMMMMMMMMMMMMMMMMMMMMMM \= M \= \kill
$x_{sum1, gts},~ x_{sum1, sum1},~ x_{sum1, sum2},~ x_{gts, sum2},~ x_{ys1,  ys2}$
    \> $=$ \> $0$ \\
$x_{sum1, ys2},~ x_{gts, ys1},~   x_{sum2, ys1}$
    \> $=$ \> $1$ 
\\[1ex]
$\pi_{sum1},~ \pi_{gts },~ \pi_{sum2}$
    \> $=$ \> $0$ \\
$\pi_{ys1 },~ \pi_{ys2 }$
    \> $=$ \> $1$ 
\\[1ex]
$c_{gts},~ c_{ys1},~ c_{ys2}$           
    \> $=$ \> $0$
\end{tabbing}
This minimal solution is not unique, though in this case the only other minimal solutions use different $\pi$ values, and denote the same clustering. Looking at just the non-zero variables in the objective function, the value is $25 \cdot x_{sum1,ys2} + 25 \cdot x_{gts,ys1} + 1 \cdot x_{sum2, ys1} = 51$. For illustrative purposes, note that objective function could be reduced by setting $x_{sum1,ys2} = 0$ (fusing $sum1$ and $ys1$), but this conflicts with other constraints. Since $x_{sum1, sum2} = 0$, we require $\pi_{sum1} = \pi_{sum2}$, but also \mbox{$\pi_{sum2} < \pi_{ys2}$}. These constraints cannot be satisfied, so a clustering that fused $sum1$ and $ys2$ would not also permit $sum1$ and $sum2$ to be fused.

We will now compare the clustering produced by our system, with the one implied by stream fusion. As described in \cite{lippmeier2013flow}, stream fusion cannot fuse a produced array into multiple consumers, or fuse operators that are not in a producer-consumer relationship. The corresponding values of the $x_{ij}$ variables are:
\begin{tabbing}
MMMMMMMMMMMMMMMMMMMMMMMMMM \= M \= \kill
$x_{gts, sum2}$
    \> $=$ \> $0$ \\
$x_{sum1, gts}, x_{sum1, sum2}, x_{ys1,  ys2}, x_{sum1, ys2}, x_{gts, ys1 }, x_{sum2, ys1}$
    \> $=$ \> $1$
\end{tabbing}
We can force this clustering to be applied in our integer linear program by adding the above equations as new constraints. Solving the resulting program then yields:
\begin{tabbing}
MMMMMMMMMMMMMMMMMMMMMMMMMM \= M \= \kill
$\pi_{sum1}, \pi_{gts }, \pi_{sum2}$
    \> $=$ \> $0$ \\
$\pi_{ys1 }, \pi_{ys2 }$
    \> $=$ \> $1$ \\
$c_{gts}, c_{ys1}, c_{ys2}$           
    \> $=$ \> $0$
\end{tabbing}
Note that although nodes $sum1$ and $sum2$ have equal $\pi$ values, they are not fused. Conversely, if two nodes have different $\pi$ values then they are never fused.

% The corresponding value of the objective function is: $25 \cdot x_{sum1, gts} + 1 \cdot x_{sum1,sum2} + 25 \cdot x_{sum1, ys2} + 25 \cdot x_{gts, ys1} + 1 \cdot x_{sum2, ys1} + 25 \cdot x_{ys1, ys2} = 102$.

% -----------------------------------------------------------------------------
\subsection{Closest Points}
The closest points benchmark is a divide-and-conquer algorithm that finds the closest pair of 2-dimensional points in an array. We first find the midpoint along the Y-axis, and filter the remaining points to those above and below the midpoint. We then recursively find the closest pair of points in the two halves, and merge the results. As the filtered points are passed directly to the recursive call, there is no further opportunity to fuse them, and our clustering is the same as returned by Megiddo's algorithm. However, our clustering generates both filtered arrays in a single loop, unlike stream fusion that requires a separate loop for each.

% -----------------------------------------------------------------------------
\subsection{QuadTree}
The QuadTree benchmark recursively builds a 2-dimensional space partitioning tree from an array of points. At each step the array of points is filtered into four 2-dimensional boxes. As with the closest points algorith, there are no further opportunities for fusing the filtered results, and our clustering is the same as Megiddo's. However, our clustering produces all four filtered results in a single loop, whereas stream fusion requires four loops.

% -----------------------------------------------------------------------------
\subsection{QuickHull}
The core of the QuickHull algorithm is shown below: given a line and an array of points, we filter the points to those above the line, and also find the point farthest from that line.

@@

Stream fusion cannot fuse the @pts'@ and @ma@ bindings because @pts'@ is referred to multiple times and thus cannot be inlined. Megiddo's algorithm also cannot fuse the two bindings because their iteration sizes are different. If the @ma@ binding was rewritten to operate over the @pts@ array instead of @pts'@, Megiddo's formulation would be able to fuse the two, and the overall program would give the same result. However, this performance behavior is counter intuitive because @pts'@ is likely to be smaller than @pts@, so in an unfused program the original version would be faster. Our system fuses both versions.
